$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-17 write_up/ms_arxiv.tex
added arxiv version of the paper

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2018

@@

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2018}

% to avoid loading the natbib package, add option nonatbib:
%     \usepackage[nonatbib]{neurips_2018}

@@

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely overlooked by existing approaches. Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

@@

The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space and observation over short intervals mean the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_2.pdf}}
	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
	\label{fig:joint}
\end{figure}

@@

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N/\bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} = N\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} two additional examples based on a multivariate normal model.

@@

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. with one chain with a lower variance). We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples. The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1}
\end{figure}

@@

\subsection{Diagnosing convergence in joint distributions: bivariate normal model}\label{sec:bivariate_normal}
To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density. In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions. By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.

@@

Next, we estimate the $R^*$ distribution which is shown in Fig. \ref{fig:bivariate}. The mean of this distribution is 1.18 and all the samples are above 1 -- indicating that the sampling distribution has not converged. By taking account of all the information in the chains,

\begin{figure}[h]
	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:bivariate}
\end{figure}

\subsection{Slow joint distribution convergence: 250-dimensional multivariate normal model}\label{sec:multivariate_normal}
In this example, we use a 250-dimensional multivariate normal target, where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \cite{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibit a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Together, these plots illustrate that the GBM's variable importance measure provides complementary information to $\hat{R}$ and ESS.

\begin{figure}[h]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5). See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy}
\end{figure}

@@

The model can be parameterised two ways, as described in \cite{vehtari2019rank}. The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
%
\begin{align*}
\theta_j &\sim \N(\mu, \tau) \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$. This can be written as,
%
\begin{align*}
\tilde{\theta}_j &\sim \N(0, 1) \\
\theta_j &= \mu + \tau \tilde{\theta}_j \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
of the distribution of these effects. In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.

We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains. Like \cite{vehtari2019rank}, we use settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased. We also used Stan with the same algorithm settings to sample from the non-centered model.

@@

In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two. The results are shown in Fig.\ref{fig:eight_schools}B. In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards. Despite this, however, the distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged.

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains. In both cases, the plots show 1000 $R^*$ samples. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:eight_schools}
\end{figure}

\section{Discussion}
If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated it. Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in samples. By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic. By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$. Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.

The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension. $R^*$ naturally includes this information in building a model capable of predicting the chain which generated each sample. It also naturally includes information about the joint distribution across all dimensions of the target. Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects. Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise. This is not a sleight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.

@@

MCMC enables inference across a large range of models encountered across the social, biological and physical sciences. Its ease of implementation, however, masks important underlying fragilities in the method. Namely, that unless the chains have converged to a truly stationary distribution, the samples generated are not faithful depictions of the posterior. In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far. $R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.

>>>>

\documentclass{article}
\pdfoutput=1
% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2018

@@

% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{nips_2018}

% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{nips_2018}

@@

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call this convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely overlooked by existing approaches. Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

@@

The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_2.pdf}}
	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
	\label{fig:joint}
\end{figure}

@@

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} and two additional examples based on a multivariate normal model.

@@

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. with one chain with a lower variance). We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples. The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates indicating that the chains have not converged.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1}
\end{figure}

@@

\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density. In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions. By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.

@@

Next, we estimate the $R^*$ distribution which is shown in Fig. \ref{fig:bivariate}. The mean of this distribution is 1.18, and all the samples are above 1, indicating that the sampling distribution has not converged. By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals. As such, we see a role for $R^*$ as an additional convergence metric.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:bivariate}
\end{figure}

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \cite{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5). See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy}
\end{figure}

@@

The model can be parameterised two ways, as described in \cite{vehtari2019rank}. The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
%
\begin{align*}
\theta_j &\sim \N(\mu, \tau), \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$. This can be written as,
%
\begin{align*}
\tilde{\theta}_j &\sim \N(0, 1), \\
\theta_j &= \mu + \tau \tilde{\theta}_j,\\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
of the distribution of these effects. In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.

We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains. Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased. We also used Stan with the same algorithm settings to sample from the non-centered model.

@@

In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two. The results are shown in Fig.\ref{fig:eight_schools}B. In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards. Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears converged.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools.pdf}}
	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains. In all cases, the plots show 1000 $R^*$ samples. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:eight_schools}
\end{figure}

\section{Discussion}
If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a sample it should be impossible to discern which chain generated it. Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in samples. By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic. By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$. Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.

The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target. $R^*$ naturally includes this information in building a model capable of predicting the chain that generated each sample. It also naturally includes information about the joint distribution across all dimensions of the target. Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects. Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise. This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.

@@

MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences. Its ease of implementation, however, masks important underlying fragilities in the method. Namely, that unless the chains have converged to a truly stationary distribution, the samples generated are not faithful depictions of the posterior. In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far. $R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
